|

### 2021-03-13 paper/v2_taylor_francis/broadband.speed.covid.Rmd
Updated descriptive clusters, paper section 4, and response to reviewers.

However, there were insufficient observations -- only $631$ speed
tests per Local Authority District (LAD) on average -- 
for each for each working hour of each working day in each LAD to profile speeds at 
that level of detail.
Spatial aggregation was also necessary because we could not follow individuals or
households and connect data points.
Therefore, we aggregate the $241,088$ individual, geolocated and time-stamped
speed-checks during the $13$ weeks of March to May inclusive for weekdays in $2020$
by each hour of the day and day of the week.
As our research aims to identify the geography of internet service resilience for
work purposes, bank holidays and the hours between midnight and $6:00$ were excluded,
as well as weekend days.
The composite week time-series thus comprise $18$ hours multiplied by $5$ weekdays or
$90$ time points per series.
The time-series were calculated for each of the $382$ LADs in the UK, standardised,
and then a *k*-means partitioning around medoids clustering algorithm
was applied using DTW.
We initially run the algorithm for $k \in \mathbb{N} \bigcap [5,15]$
and used cluster validity indices (CVIs) to pick the optimal solution of $k = 13$.
Following @sardatime the majority vote for the following CVIs was used: Silhouette
(max), Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min), Modified
Davies-Bouldin (DB*, min), Dunn (max), COP (min).

@@

<!-- metrics -->
In Section [4.1](#sec:4.1), we review the temporal profile of upload speed
by hour of the day and day of the composite week for each cluster.
Since the quality and reliability of internet services vary in time and space due
to both supply and demand-side influences, we also use a number of different measures
to describe experienced upload speeds per cluster.
These include: i) mean, experienced connection speed,
ii) standard deviation or the amount of fluctuation from the mean, and
iii) the variation in speeds during the new morning peak of testing when working
from home is more likely to take place.
We take account of all three measurements
in order to determine how resilient broadband speeds are
as experienced in different parts of the UK during a time of extreme demand.

@@

The temporal profiles of the local authority clusters have been summarised in
Figures \ref{UpClusterL} and \ref{UpClusterS} and Table \ref{up.cluster.descr}.
The graphs show a composite profile of mean upload speeds per hour
per day for each cluster, with the largest five clusters, 
in terms of the LAD membership and population, in Figure \ref{UpClusterL},
and the next six in Figure \ref{UpClusterS}.
These figures and table provide a comprehensive overview of the quality and
reliability of experienced broadband in different parts of the UK.

@@

stargazer(clustersLADup[1:7],
          summary = F,
          font.size = "footnotesize",
          type = "latex",
          header = F,
          digit.separator = ",",
          digit.separate = c(3,3),
          covariate.labels=c("Cluster",
                             "N. of LADs",
                             "LAD population",
                             "mean speed",
                             "SD speed",
                             "mean AM speed",
                             "mean PM speed"),
                             #"AM_slower",
                             #"SDtoMean"),
          rownames = F,
          no.space=TRUE,
          column.sep.width = "0pt",
          notes = "Note: All speed measures are upload speeds",
          title = "Upload speed cluster characteristics\\label{up.cluster.descr}")
```
The second largest cluster, comprising $126$ local authorities and over $20$ million
people, is cluster $13$, which has the slowest aggregate mean upload speed,
and the second highest ratio of the standard deviation to the mean.
This suggests that those living in local authorities in this cluster experienced
some of the lowest quality broadband services in terms of upload speeds in the UK.
However, Figure \ref{UpClusterL} shows that the high standard deviation,
which is one indication of unreliable internet, did not disproportionately affect
the morning peak from $9:00$-$10:59$, when upload speeds were, on average,
only $2.6$% slower than in the evening peak period between $19:00$ and $20:59$,
when entertainment purposes are likely to be using the most bandwidth.
In comparison, the five LADs in cluster $1$, home to $343$ thousand people, not only
experience the second slowest mean upload speeds and the highest ratio of standard
deviation to the mean, but are also much more affected during the morning peak.

Meanwhile, those living in the largest cluster -- $6$, with $155$ LADs and $29.5$
million people -- experienced aggregate mean upload speeds of about $1$Mb/s faster
than those in cluster $13$, but still lower than the other three large clusters
and most of the smaller clusters, suggesting a middling quality of service.
The temporal profile for cluster $6$ in Figure \ref{UpClusterL} shows that upload
speeds are highest at $6:00$ on a Monday morning and between $23:00$ and midnight 
on Wednesday and Thursday, but tend to be lower during the working day.
Furthermore, experienced mean upload speeds in the morning peak are $4.4$% lower
than in the evening peak -- a greater, more noticeable change than any of the other
large clusters experience, suggesting poorer reliability during the working day.
This difference is less, however than any of the clusters
included in Figure \ref{UpClusterS}.

Clusters $8$ and $12$ also have mean upload speeds under $10$Mb/s, but higher than
clusters $1$, $13$, and $6$. However, mean upload speeds are much lower between
$9:00-10:59$ than between $19:00-20:59$ in cluster $8$, but slightly faster in the
morning than in the evening in cluster $12$.
Indeed, cluster $12$ is the only cluster to experience higher speeds in the morning
peak compared to the evening, and thus the only cluster where the temporal profile
of internet use is closer to what might have been expected pre-pandemic.
Among the other clusters, however, the reliability of internet services
during the working day still varies considerably.
Interpreting this variation from the large spikes and dips shown on
Figures \ref{UpClusterL} and \ref{UpClusterS} is difficult,
but the statistics in Table \ref{up.cluster.descr} show that clusters $9$ and $11$
have the most reliable internet services.
The ratio of standard deviation to mean in both these clusters is lowest,
and speeds are only about 2% slower in the morning than the evening.
Mean speeds are also higher than in any other cluster, excluding cluster $2$,
where measures of reliability suggest poorer performance.

Thus, broadband services in clusters $9$ and $11$, 
home to over twelve million people, performed the best during the study period, 
in terms of both quality and reliability.
In Figure \ref{UpClusterL}, cluster $11$ shows more noticeable peaks and troughs,
with the lowest points occurring, on average, between $6:00-7:00$ on Monday morning, 
$14:00-15:00$ on Friday, and $16:00-17:00$ on Thursday, not the workday peak. 
Furthermore, these slow periods offer better speeds than the average hourly
profile of cluster $13$. Clusters $3$, $7$ and $10$ also have relatively high 
mean upload speeds. Clusters $3$ and $10$ pair high mean speeds with
low standard deviations relative to the mean speeds, suggesting reliability and
resilience, as well as quality broadband services. Cluster $7$ has a higher ratio of
standard deviation to mean, but there is less difference in average speeds between
the morning and evening peaks than in clusters $3$ and $10$.

In summary, LADs in clusters $9$ and $11$ experienced resilient
broadband internet that could support high levels of telecommuting.
Those in clusters $2$, $3$, $7$, and $10$ also experience higher than average
mean speeds and rank middle to high on measures of service reliability.
These LADs are *not* on the wrong side of the first level digital divide,
but how likely are they to be able to take advantage of their resilient ICT
infrastructure and services?
Meanwhile, cluster $6$ is not only the largest in terms of number of LADs
and population, it has the closest mean upload speed to the pre-clustered average
for the whole sample.
As well as average quality internet services, those in cluster $6$ also experience
average reliability for work purposes, ranking fifth behind the four other clusters
with populations over one million, but ahead of the smaller clusters.
Clusters $8$ and $12$ are also close to average mean upload speeds, but show very
different patterns in terms of reliability, whilst clusters $1$ and $13$ appear to
suffer most from a lack of quality internet services, with slow speeds and
high standard deviations.
With those in cluster $1$ more likely to experience that poor reliability during the
morning peak, is this first level digital divide occurring in areas where few are
occupationally able to telecommute anyway, and what are the implications for
economic resilience?

@@

Using an auxiliary multinomial logit regression, we test whether the clusters that
have higher mean speeds and more reliable services consist of LADs that are more
urban, affluent, and / or more likely to benefit from a choice of high quality
internet services.
We also estimate which of our clusters are more likely to have a higher proportion of
occupations where the nature of the work enables telecommuting.
The results of the auxiliary regression are presented in Table \ref{aux}.
The dependent variable is the LAD cluster membership as described in the
[methods and data](#sec:3) section
and equation \ref{eq1}. Each column represents a different cluster.
The reference case is cluster $4$, which includes only the local authority of
Hambleton in North Yorkshire, a rural area of just over ninety thousand people.
Mean, experienced upload speeds in cluster $4$ (see Table \ref{up.cluster.descr})
are between the average speeds for the $13$ clusters ($9.9$Mb/s) and the
pre-clustered average for the whole sample ($9.3$Mb/s).
However, the standard deviation for cluster $4$ and the difference between average
speeds in the morning compared to the evening peak periods are indications of
worse reliability than many of the other clusters. Hence, the results in Table
\ref{aux} should be seen as relative rather than absolute probabilities.

First, we control for the number of speed tests run per cluster inhabitant
between $9:00$-$10:59$ as well as the share of fast Virgin Media internet
connections. Regarding the former, we expect people in LADs with more unreliable
connections to test their internet speeds more often, and the results validate our
priors. 
Those in clusters $2$, $3$, $9$ and $11$ benefit from a higher proportion of
Virgin connections, which is an indication that people in these clusters are
more likely to live in urban areas, with more choice of broadband services.
In other words, they are more likely to be on the right side of the first level,
infrastructure-based digital divide, as we expected from the analysis in Section
[4.1](#sec:4.1).

We employ distance from London and from the nearest metropolitan area
(including London) as two variables depicting peripherality, urban structure and,
potentially, first level digital divides.
The broadband speed tests run in the authorities in cluster $3$ are more likely to
be taking place close to London than those run in any of the other clusters, and
two of the four authorities in cluster $3$ are the London commuter towns of Harlow
and Luton.
However, even though London was also included in the variable calculating distance
from the centre of one of either the ten largest metropolitan areas in England,
or Glasgow or Cardiff, tests run in cluster $3$ are likely to be furthest away.
Thus, it is important to consider the membership of each cluster as well as the
regression results. Corby and Eastbourne, the other two authorities in cluster $3$
are large, accessible towns, although not part of metropolitan areas.

Meanwhile, LADs in cluster $7$ are most likely to be near the centre of a large
metropolitan area, even though the four local authorities of cluster $7$ include
no central urban boroughs and only one LAD that is part of a metropolitan area of
governance - Tameside in Greater Manchester. This may explain why those in cluster
$7$ are the second least likely to be served by Virgin Media.
It is also a demonstration of the complexity of both experienced broadband 
upload speeds as captured by time-series clustering, and the geography of 
first level digital divides as a product of quality as well as availability.
In comparison, cluster $1$, which our analysis suggests lacks broadband resilience, 
contains five, mainly rural authorities, but they are closer to a metropolitan area 
than authorities in cluster $3$, although furthest from London, perhaps because 
they are scattered around the country -- see Figure \ref{map.up.clusters}.

Internet resilience is also more nuanced than our -- arguably crude
-- dummy variable depicting the North-South economic divide, which assigns $1$ to
LADs located in Greater London, Southeast, Southwest and East of England regions,
and $0$ to the rest.
The authorities most likely to be in the South are those in cluster $13$,
which was identified as having the slowest mean upload speeds of any of the clusters,
and thus a low level of service.
However, cluster $13$ includes some rural, remote areas of the country, such as
Northwest Scotland, Cornwall and Powys in Wales as shown on Figure
\ref{map.up.clusters}.
It also includes major metropolitan centres in the North -- Liverpool, Newcastle --
and South -- Bristol, and nine (of $32$) London Boroughs.
There are also plenty of Southern home county and suburban areas.
The standard deviation measure for cluster $13$ is high, but speed variation is
low during the morning peak, suggesting that the estimates for reliability are
inconsistent. Considering that this is one of the largest clusters, and thus the
averages incorporate more noise than some of the smaller clusters,
it may be that the LADs in this cluster do not all suffer equally from
a first level digital divide.

Yet we need to consider the results for other variables in Table \ref{aux} to better
determine whether clusters $1$ and $13$, which appear to suffer most from poor
quality internet services, are also more likely to have a low skilled workforce,
less able to benefit from telecommuting.
Cluster $1$ has the lowest proportion of educated people,
and the lowest earnings, despite recording the highest proportion of managerial and
professional jobs, and the second highest proportion of tech jobs.
Rural areas such as those in cluster $1$ are home to many older, retired people 
[@blank2018local], which might explain these results or perhaps, 
as these figures are relative, we could note that cluster $1$ also has a greater 
proportion of skilled trades than other clusters. 
In any case, it appears that the first level digital divide 
reinforces other inequities in cluster $1$, where the even slower than average 
morning upload speeds shown in Table \ref{up.cluster.descr} suggest that internet
users were more active during the working day. Meanwhile, those in cluster $13$ are 
more likely to earn more and have a better education despite poor internet services.
Cluster $13$ also has the most businesses per inhabitant, but is somewhere in the
middle in terms of job density. If this is an indication of a high number of SMEs,
it could explain the variable internet quality, considering that small businesses
have not been seen as the most valuable customers for higher speed broadband services
if they are not located near residential customers [@ofcom2016].

Cluster $6$ is the largest cluster, and thus, like cluster $13$, there is more
noise within the averages we use to measure internet quality and reliability.
Our results indicate average mean speeds, and Table \ref{aux} shows that
cluster $6$ also falls towards the middle of the clusters on many of the
socioeconomic variables. It ranks third or fourth out of the eleven in terms of
the likelihood of having a higher proportion of managerial, professional, and
tech jobs, as well as higher earnings, but is fourth from bottom for educational
attainment. The LADs in cluster $6$ are also diverse, with few truly remote areas,
but urban areas throughout England, including Birmingham, Leeds, Sheffield,
twelve London Boroughs, and many suburban areas and smaller cities like Oxford
and Cambridge. The capital cities of the other UK nations, Belfast, Cardiff and
Edinburgh, are also in this cluster, suggesting perhaps that the lower level of
reliability discussed in Section [4.1](#sec:4.1) may be due to increased demand,
e.g. for telecommuting, despite lacking the most resilient internet connections.

In contrast, Clusters $9$ and $11$ enjoy resilient internet connections, 
but LADs in Cluster $9$ are more likely to host highly educated individuals 
with higher earnings than cluster $11$, which has a negative coefficient for 
the NVQ4 variable. Cluster $11$ has fewer businesses per inhabitant and the 
second highest job density and cluster $9$ the second lowest.
These coefficients might indicate that resilient broadband infrastructure
generates higher returns for those in cluster $9$, where slightly more slowdown in
the AM peak was detected -- see Table \ref{up.cluster.descr}.
LADs in cluster $9$ are less likely to be in the South and those in cluster $11$
more likely, although the coefficient for cluster $9$ might be skewed by
the presence of the Scottish cities of Glasgow and Aberdeen. Scotland has a
different economic profile than England. Still, both clusters consist of larger
LADs in terms of population, including districts within five of England's ten
largest cities, and a number of other stand-alone urban areas and large market towns 
-- see Appendix [1](#appendix1).
These urban locations are on the right side of the first level digital divide,
and the regression results suggest that internet resilience is supporting a wide
range of small and large urban economies.

LADs in clusters $2$, $3$, $7$, $10$ are also on the right side of the first level
digital divide, experiencing higher than average mean upload speeds and ranking 
high to middle on measures of service reliability.
LADs in clusters $2$ and $10$ are also more likely to have highly skilled
workers to take advantage of working from home opportunities.
Cluster $2$ is comprised of just two LADs, with the lowest job density of any
cluster. Cluster $10$ is comprised of four LADs, including two
peripheral suburban areas of Birmingham, the East London Borough of Newham,
and Dundee. Suburbs are considered the most likely urban form in which telecommuters
live [@e2018does], and Dundee has a reputation for tech startups [@technation2017].
LADs in cluster $7$, which include the suburbs South of Belfast, a suburban
district of Greater Manchester, and some other Midlands and North Yorkshire towns 
and villages have the highest earnings, and thus may also be making the
most of their resilient internet. In contrast, LADs in cluster $3$ are associated
more with lower skills and have the highest proportion of individuals in machine
operation jobs. Arguably, LADs in clusters $3$ benefit the least from their
resilient internet infrastructure in an era when working from home became a vehicle
for economic resilience.

Finally, clusters $8$ and $12$ consist of LADs enjoying close to average
mean upload speeds, but opposing patterns of internet reliability during the morning
peak. LADs in cluster $8$ are likely to host more individuals employed in tech and
administrative occupations than any other cluster, as well as many in managerial
occupations, whilst the opposite applies to skilled trade and leisure occupations.
These LADs are characterised by a very small positive likelihood of more educated
residents, but are not significantly likely to earn more than other clusters.
Including suburbs near Leicester, Cardiff, and Birmingham and south of London, 
the LADs in cluster $8$ seem to likely to have the skills and occupations that 
would benefit from quality internet services, but suffer from poor internet 
resilience and reliability.

In comparison, LADs in cluster $12$ are less likely to be able to benefit from
quality internet services if they had them, with fewer individuals achieving
NVQ4 or better and lower levels of occupations that would benefit from homeworking.
Yet LADs in this cluster benefit from the second highest level of earnings, and
have the largest stock of businesses per inhabitant and the highest job density.
This density of businesses could be why cluster $12$ is the only cluster with higher
speeds in the morning peak during the study period. If people are at home, business
premises might well have been abandoned. Cluster $12$ is home to $1.5$ million people
spread across twelve LADs, from London to Wales as listed in Appendix [1](#appendix1). 
This spatial diversity demonstrates that the temporal clustering of internet
resilience is not necessarily spatially dependent, and digital divides do not
necessarily overlap with economic ones.

# Conclusions {#sec:5}

>>>>

However, there were insufficient observations -- only $631$ speed
tests per Local Authority District (LAD) on average -- 
for each for each working hour of each working day in each LAD to profile speeds at 
that level of detail.
Spatial aggregation was also necessary because we could not follow individuals or
households and connect data points.
Therefore, we aggregate the $241,088$ individual, geolocated and time-stamped
speed-checks during the $13$ weeks of March to May inclusive for weekdays in $2020$
by each hour of the day and day of the week.
As our research aims to identify the geography of internet service resilience for
work purposes, bank holidays and the hours between midnight and $6:00$ were excluded,
as well as weekend days.
The composite week time-series thus comprise $18$ hours multiplied by $5$ weekdays or
$90$ time points per series.
The time-series were calculated for each of the $382$ LADs in the UK, standardised,
and then a *k*-means partitioning around medoids clustering algorithm
was applied using DTW.
We initially run the algorithm for $k \in \mathbb{N} \bigcap [5,15]$
and used cluster validity indices (CVIs) to pick the optimal solution of $k = 13$.
**ET edit this bit for rationale k = 9**
Following @sardatime the majority vote for the following CVIs was used: Silhouette
(max), Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min), Modified
Davies-Bouldin (DB*, min), Dunn (max), COP (min).

@@

<!-- metrics -->
In Section [4.1](#sec:4.1), we review the temporal profile of upload speed
by hour of the day and day of the composite week for each of the clusters.
Since the quality and reliability of internet services vary in time and space due
to both supply and demand-side influences, we also use a number of different measures
to describe experienced upload speeds per cluster.
These include: i) mean, experienced connection speed,
ii) standard deviation or the amount of fluctuation from the mean, and
iii) the variation in speeds during the new morning peak of testing when working
from home is more likely to take place.
We take account of all three measurements
in order to determine how resilient broadband speeds are
as experienced in different parts of the UK during a time of extreme demand.

@@

The temporal profiles of the local authority clusters have been summarised in
Figure \ref{UpCluster} and Table \ref{up.cluster.descr}, to provide an overview of 
the quality and reliability of experienced broadband in different parts of the UK.
Figure \ref{UpCluster} shows a composite profile of mean upload speeds per hour
per day for each of the largest six clusters, in terms of the LAD membership and 
population.

@@

stargazer(clustersLADup[1:7],
          summary = F,
          font.size = "footnotesize",
          type = "latex",
          header = F,
          digit.separator = ",",
          digit.separate = c(3,3),
          covariate.labels=c("Cluster",
                             "N. of LADs",
                             "LAD population",
                             "mean speed",
                             "SD speed",
                             "mean AM speed",
                             "mean PM speed"),
                             #"AM_slower",
                             #"SDtoMean"),
          rownames = F,
          no.space=TRUE,
          column.sep.width = "0pt",
          notes = "Note: All speed measures are upload speeds",
          title = "Upload speed cluster characteristics\\label{up.cluster.descr}")
```
The largest cluster, comprising $229$ local authorities and over $40$ million
people, is cluster $6$, which has the slowest aggregate mean upload speed of any
cluster, and the highest ratio of the standard deviation to the mean.
This suggests that those living in local authorities in this cluster experienced
some of the lowest quality broadband services in terms of upload speeds and
reliability in the UK. However, as shown in Figure \ref{map.up.clusters}, some of the
most rural areas of the UK are included in this cluster. If these areas
suffer most from first level digital divides as described in the literature review,
the low speeds in rural areas might be pulling down the averages in other areas in
this largest cluster. Also, since the areas are clustered by their temporal profile
across the working week, the graph in Figure \ref{UpCluster} indicates that speeds
in cluster $6$ are some of the more reliable. Table \ref{up.cluster.descr} confirms
that upload speeds in cluster $6$ during the morning peak from $9:00$-$10:59$ were, 
on average, only $4.5$% slower than in the evening peak period between $19:00$ 
and $20:59$, when entertainment purposes are likely to be using the most bandwidth.

Those living in the second largest cluster -- $9$, with $115$ LADs and $21.5$
million people -- experienced aggregate mean upload speeds of more than $1.5$Mb/s 
faster than those in cluster $6$, a lower ratio of standard deviation to mean, and
almost the same upload speeds in the AM Peak as the PM Peak.
Indeed, the temporal profile for cluster $9$ in Figure \ref{UpCluster}, like that
for cluster $6$, is fairly flat. Whilst it is likely that the large numbers of 
tests being performed by the large populations in these two clusters of LADs result 
in less varied averages to show in the graph, the method of shape-based clustering
suggests that these results indicate that the majority of the population of the 
UK experience less temporal variation, and so belong in these two clusters.
The results also confirm that those in cluster $9$ experienced a consistently better 
service during lockdown than those in cluster $6$, in terms of both average speed 
and reliability.

Of the smaller clusters, cluster $7$ experiences quite similar mean speeds to 
cluster $9$, and a not too dissimilar standard deviation, but mean speeds in the 
morning peak are almost $10$% lower than the evening peak.
Cluster $3$ boasts speeds almost $1$Mb/s higher again, but suffers from even greater
slowdown in the morning peak. In comparison, clusters $2$ and $5$, home to a 
little less than half a million people, experience not only above average mean 
speeds, but also faster speeds in the morning peak compared to the evening.
Thus, in the five LADs in these two clusters, the temporal profile of internet use 
may be closer to what might have been expected pre-pandemic.
In comparison, the $15$ LADs in clusters $1$ and $8$, home to over $1.4$ million 
people, not only experience fairly average upload speeds and high ratios of 
standard deviation to the mean, but also experience much lower speeds during 
the morning peak than the evening peak. From the large spikes and dips shown on
Figure \ref{UpCluster}, it is likely that this poor reliability or consistency of 
morning internet speeds was more noticeable to those in cluster $1$, whilst those 
in cluster $8$ experienced the lack of reliable service as a problem throughout 
the day.

In summary, the reliability of internet services during the working day appears to
have altered for the vast majority of locations due to increased use from residents 
told to stay at home, and, if they were fortunate, work from home.
Yet this experience has been different in different locations. In particular, 
LADs in cluster $9$ experienced higher speeds and more resilient
broadband internet than those in the largest cluster, $6$.
Those in clusters $3$ and $7$ also experience higher mean speeds and better service 
reliability, and can be more confident that they are on the right side of the first 
level digital divide, and that their more resilient ICT infrastructure and services
can robustly support higher levels of telecommuting.
On the other hand, whilst high levels of telecommuting may be the cause of morning 
speeds over 15% lower than those in the evening, those in clusters $1$ and $8$ 
also experience lower mean speeds than clusters $3$, $7$, and $9$, but not as low as 
cluster $6$. Whilst the differences may not be large, the lower upload speeds in 
cluster $6$ or the poor reliability in clusters $1$ and $8$ could still have had 
implications for economic resilience if combined with other types of digital divide.
This potential will be explored in the next section.

@@

Using an auxiliary multinomial logit regression, we test whether the clusters that
have higher mean speeds and more reliable services do indeed consist of LADs that 
are more urban or closer to major urban areas and are more likely to benefit from 
a choice of high quality internet services. This part of the regression aims to
confirm any first level digital divides.
Next, to better understand how the clusters fare in relation to the second level
digital divide, we consider which LADs in which clusters have a higher proportion of 
occupations where the nature of the work and the skills that occupation employs 
enable telecommuting. 
Finally, we consider the rate of return on internet use, or third level digital 
divides, by reviewing which clusters have the highest earnings and job density,
and also which clusters experienced the highest share of population furloughed 
during this period in the pandemic.

The results of the auxiliary regression are presented in Table \ref{aux}.
The dependent variable is the LAD cluster membership as described in the
[methods and data](#sec:3) section and equation \ref{eq1}. 
Each column represents a different cluster.
The reference case is cluster $4$, which includes only the local authority of
Hambleton in North Yorkshire, a rural area of just over ninety thousand people.
Mean, experienced upload speed in cluster $4$ (see Table \ref{up.cluster.descr})
is close to the pre-clustered average for the whole sample ($9.3$Mb/s).
However, the standard deviation for cluster $4$ and the difference between average
speeds in the morning compared to the evening peak periods are indications of
worse reliability than many of the other clusters. Hence, the results in Table
\ref{aux} should be seen as relative rather than absolute probabilities.

First, we control for the number of speed tests run per cluster inhabitant
between $9:00$-$10:59$ as well as the share of fast Virgin Media internet
connections. Regarding the former, we expect people in LADs with more unreliable
connections to test their internet speeds more often, and the results show that 
those in cluster $6$ are by far the most concerned about reliability at that time 
of day. Meanwhile, those in clusters $2$, $3$, and $9$ benefit from a higher 
proportion of Virgin connections, which is an indication that people in these 
clusters are more likely to live in urban areas, with more profitable broadband 
markets, and a better choice of broadband services. 
We also employ distance from London and from the nearest metropolitan area
(including London) as two variables depicting peripherality and urban structure.
However, whilst significant, the pattern of coefficients of these variables is 
inconclusive in confirmed first level digital divides. 
Even though London was one of the ten largest metropolitan areas in 
England, which, along with Glasgow and Cardiff, were identified to calculate the 
variable estimating the impact of distance from the centre of a metropolitan area, 
the coefficients for the two variables are mostly small and the signs for some are 
opposite.

When looking at the constituent authorities (see full list of LADs in 
Appendix [1](#appendix1) as well as Figure \ref{map.up.clusters}), 
however, Cluster $9$ is clearly more urban, including $13$ of $32$ London Boroughs, 
five of the seven constitutent LADs of the West Midlands conurbation, 
nine of the ten boroughs of Greater Manchester, and five of nine other 
large metropolitan areas coded in for the 'distance to nearest met area' variable. 
Cluster $9$ also includes the other main cities in the East Midlands, Leicester 
and Derby, as well as smaller cities known for their knowledge economy, such as 
Oxford and Milton Keynes. 
However, as previously mentioned, there is more noise within the variables for the 
two largest clusters than the smaller ones, and cluster $6$ also contains many urban 
areas. These include $16$ London boroughs, Birmingham City, Bolton in Greater 
Manchester, Leeds, Liverpool, Newcastle and Bristol, as well as small cities with 
knowledge economies such as Cambridge, Edinburgh, and Reading and its neighbours in 
the high-tech agglomeration of the Thames Valley. 
And yet cluster $6$ also includes some of the most rural areas in the country, 
showing how important it is to interpretation to consider the membership of the 
cluster as well as the regression results.

The coefficients in Table \ref{aux} demonstrate that those in cluster $6$ were more 
likely to hold managerial, tech and professional jobs than those in cluster $9$, and 
were the least likely to be furloughed of any cluster. 
Thus, although speeds were slow and not as reliable during the morning peak 
as in cluster $9$, the skills that enabled telecommuting also enabled greater 
returns from doing so, despite lower than average earnings.
Cluster $6$ also has the second highest job density, with more jobs to keep.
Likewise, those in LADs in clusters $1$ and $8$ had the highest and second highest 
proportions of residents working in tech or managerial occupations despite suffering 
from lower speeds and unreliable services, another mismatch between first and 
second level digital divides. Yet unlike cluster $6$ and despite occupations with 
better digital skills, cluster $8$, with its five peripheral suburban areas in the 
Midlands, Scotland and south of London, as well as rural West Devon saw the third 
highest proportion of its working population furloughed. It also had the second 
lowest average earnings in 2019, suggesting that its population was more likely to
be on the wrong side of the third level digital divide before, as well as during 
the pandemic. In contrast, the nine LADs of cluster $1$, which include more rural 
areas, plus Westminster in central London had the lowest earnings of any cluster, 
but also lowest level of furlough after cluster $6$.

The regression results for cluster $7$ also reveal the complexity of intersecting 
and diverging digital divides. Along with $3$ and $9$, the analysis in 
Section [4.1](#sec:4.1) suggested this cluster was on the 'right' side of any first
level digital divides, yet the regression shows that it has the lowest 
proportion of Virgin connections. It is furthest from a metropolitan area, 
but closest to London. The five LADs in cluster $7$ are all suburban, making it 
sensible that the cluster is in the group with better infrastructure, but none are 
central urban boroughs and, along with a London suburb, Leeds suburb, Manchester 
suburb, and West Midlands suburb, the fifth LAD is outside Belfast. 
Belfast was not included among the metropolitan areas, and whilst there has been 
substantial investment in high speed broadband in Northern Ireland in recent years, 
including by Virgin, take-up may still lag behind historic networks in other 
urban markets.

Cluster $7$ has the lowest proportion of residents currently employed in 
professional occupations, and among the lowest proportions in managerial or 
technical occupations. This suggests potentially a low level of skills to take 
advantage of the quality internet service available, a second level digital divide. 
Cluster $7$ also has the lowest job density. However, with the largest proportion 
of skilled tradespeople, and the second highest number of businesses per inhabitant, 
those in cluster $7$ ranked second in terms of earnings in 2019. On the other hand, 
the lack of some of the occupations most likely to telecommute during lockdown 
may have contributed both to a substantial proportion being furloughed in 2020, 
and even to the relative reliability of broadband speeds compared to many of the 
other smaller clusters.

Cluster $3$, meanwhile, had the greatest proportion of its workforce furloughed of
any cluster, despite having the highest job density and the second highest average 
speeds. Ranking in the middle for most of the occupations estimated, at least some 
of its population should have the skills to work from home. The geography would 
suggest economic advantage as well, with eight of its twelve LADs in the South 
of England, including two London Boroughs, four suburban areas north of London, 
and the two tightly bounded urban areas of Eastbourne and Ipswich -- a greater 
proportion than any other cluster. 
Yet geographic position and job density do not appear to equate to particularly high 
average earnings in 2019, nor did they protect residents from furlough in 2020. 
In contrast, cluster $5$, comprised of three suburban districts in the Northwest, 
Midlands, and the London green belt, had the highest average earnings in 2019, 
despite low levels of residents working in managerial, tech and professional 
occupations. Suburbs are considered the most likely urban form in which 
telecommuters live [@e2018does], yet the contrast between these small, mainly 
suburban clusters -- $7$, $3$, and $5$, suggest that they are not all equally
economically resilient in a pandemic.

Cluster $2$, which, like cluster $5$, had faster morning broadband speeds than 
evening speeds, as shown in Table \ref{up.cluster.descr} contains two remote LADs, 
the Isles of Scilly and Northeast Lincolnshire.
Whilst the low demand may be due to the least residents in tech occupations, it may
also be due to high numbers of retired people. Rural areas such as those in cluster 
$2$ are often home to many older, retired people [@blank2018local], which may also 
explain some of the contradictions in cluster $1$ with its low levels of furlough
and low earnings, whilst the presence of Westminster in that cluster might be why
there are the highest levels of tech and managerial occupations.

In summary, the regression results indicate the complexity of measuring second and 
third level digital divides within spatial aggregations where the geography of 
first level digital divides has been captured using time-series clustering of
experienced broadband upload speeds as a product of reliability not just
availability. Whilst the differences in mean speeds between clusters were not large,
the temporal clustering of internet resilience showed much greater variation, 
and was not spatially dependent upon distance to large urban areas, or relative
location. Internet resilience supports a wide range of small and large urban 
economies, but also fails, or at least frustrates a wide range of other urban 
economies, as demonstrated by the number of tests per capita run in cluster $6$.
Yet our analysis of the likelihood of furlough within the population, especially in 
the smaller suburban clusters $7$, $3$, $5$, and also $8$ shows that digital divides 
were as likely to diverge as intersect during the pandemic, and did not necessarily 
overlap with prior economic divides. With a much greater share of the population 
likely to continue to work from home in the future, and with changing attitudes 
towards residential locations, our analysis suggests that first level digital divides
should be seen as a function of reliability as well as availability, of upload 
speeds as well as download speeds, and of greater spatial variation than might
have previously been considered.

# Discussion and Conclusions {#sec:5}

This paper offers a new perspective on telecommuting from the viewpoint of the
complex web of digital divides. We employ novel data regarding
experienced upload speeds and time-series clustering methods,
a family of unsupervised machine learning techniques which are rarely utilised
in geographical research. Fast, reliable internet connections are
necessary for the population to be able to work from home.
Although not every place hosts individuals in occupations which allow for
telecommuting nor with the necessary skills to effectively use the internet to
telecommute, this paper raises the issue that places may depend upon good internet 
reliability as well as connectivity to achieve economic resilience in a period like 
the current pandemic when internet resilience is so vital.

Our analysis demonstrated that the temporal profiles of seven of our
nine clusters had slower upload speeds in the morning than in the evening.
The opposite is likely to have been the norm prior to the pandemic, as level of
demand and bandwidth management is the most common cause of temporal variation in
experienced speeds, and why evening download speeds, rather than daytime upload
speeds, have been used to benchmark the performance of internet services.
Thus, the new patterns can be taken as evidence of widespread telecommuting
and other daytime internet use which changed the temporal profile of
internet activity throughout the UK, not just in areas with more digital industry
or better skills. Furthermore, upload speeds have not previously been seen as 
integral to universal service, considering there has never before been such extreme 
demand for telecommuting and operations such as video calls. Yet those in the 
largest cluster, $6$, were clearly frustrated by lower upload speeds. 
Their speeds were also less reliable, perhaps because they had higher levels of
demand than the second largest cluster, $9$, where many more employees were
furloughed.

Almost all of the largest urban areas in the UK, as well as many smaller cities, 
were split between the largest two clusters -- $6$ and $9$ -- and both contain LADs 
that are centres of the knowledge economy, from Cambridge in the former to Oxford 
in the latter, or have high concentrations of digital businesses, like Reading in 
the former and Milton Keynes in the latter [@technation2017].
Yet there are both intersections and divergences of digital and economic divides.
LADs in cluster $9$ were able to benefit both from reliable internet connections and
populations which were familiar with working from home and could capitalise on their 
digital infrastructure, although overall, they appeared to have a lower rate of
return in the pandemic, with far greater numbers furloughed. 
Is this an indication of both historic economic division, as well as the third level 
digital divide in cities in the North and Midlands unable to capitalise on their 
digital infrastructure?
Meanwhile, did the greater presence in London and the South of LADs in cluster $6$,
as well as the inclusion of many rural areas less impacted by the pandemic help
those in this cluster be more resilient and gain greater returns, despite using 
less reliable internet services?
The nuanced picture we gained through our analysis of the UK case study suggests
that in cluster $1$ as well as $6$, being on the right side of the second level
digital divide had a greater impact on economic resilience and therefore the third
level digital divide, than having quality internet connectivity.
